A performance assessment of evolutionary algorithms in a volunteer computing system

### 2016-03-20 PPSN2016/paper.Rnw
Fixes latex
Fixes URL and typos

\begin{abstract}
In a volunteer distributed computing system, users run a program on
their own machine to contribute to a common effort. If the program is
embedded in a web page, collaboration is easy, but also ephemeral. In
this paper, we analyse a volunteer evolutionary computing system called
NodIO, trying to find out a few rules that encourage volunteer
participation and how to make the most of their contributions to the
evolutionary algorithm. We will show different measures of
participation and contribution to the algorithm and how different
volunteer usage patterns and tweaks in the algorithm contribute to
%Pablo: añadiría un ", such as...," con un ejemplo entre "algorithm" y "contribute"
improvements and leveraging of these contributions. 
\end{abstract}

@@

The world wide web not only provides a platform for content
distribution, but also an increasingly reliable and high-performance
platform for running distributed applications. This is based in two
key facts: the Javascript virtual machine every browser runs and a
simplified standard interface for interacting with servers.
% These are "things", not "facts". As facts would be... "every browser runs a Javascript virtual machine and that there is a simplified standard interface for interacting with servers (?)
Thus
creating a distributed computing experiment is just a matter of making
a JavaScript application interchange information with a server, called
REST. From the point of view of the programmer, this involves
relatively common skills and no special libraries, since that
interface is built in the browser, and a simple application that
responds to those requests on the server side; both involve just a few
dozens lines of code additionally to whatever business logic your
application has. But, more importantly and from the point of view of
the user, that application can be run by simply visiting a web page.

@@

Volunteer computing involves a user running a program voluntarily
and, as such, has been deployed in many different ways from the
beginning of the Internet, starting with the SETI@home framework for
processing extraterrestrial signals \cite{david-seti:home}. However
the dual introduction of JavaScript as a universal language for the
browser and the browser itself as an ubiquitous web and Internet client has
made this combination the most popular for volunteer computing
frameworks such as the one we are using here, and whose first version
was described in \cite{DBLP:conf/gecco/GuervosG15}.

@@

In general, those papers go no further than a proof of concept,
attempting to gauge how many users join the effort and also how long
would it take for a concrete amount of users to reach the solution. In
general, they do not try to assess the performance of the algorithm
itself in the precise conditions they find in a volunteer computing
system, although papers such as the one by Laredo et
al. \cite{churn08:ijhpsa}, using models, try to find out the resulting
performance when the users are changing continously according to a
Weibull distribution. In the case of Klein and Spector \cite{unwitting-ec}, the algorithm
is actually run on the server, using the browser mainly for fitness
evaluation so no actual contribution to the evaluation of the
evolutionary algorithm itself is made.

@@

All experiments are announced in the same way: we use social networks,
mainly Twitter, to announce the start of a new experiment. LinkedIn,
Facebook, and also private groups in Telegram and WhatsApp have been
used for announcements too. We also encourage friends and followers to
put an original post for announcing it, instead of just sharing, in
the way allowed by the platform, the original post. During the running
of the experiments, users are engaged, questions answered, and all
kind of explanations given if requested. In general, announcements
have been made in Spanish, but in some cases English was used
too. This makes for a certainly unrealiable experimentation framework,
but it is a realistic one. The steps in the design of the evolutionary
algorithm will be explained next.

@@

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=3>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Ranked and normalized number of contributions (PUTs) per IP,
  plotted in a log $y$ scale.\label{fig:zipf}}
\label{fig:zipf}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP request) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that are slightly superior to the rest, specially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

%
\begin{table}
  \caption{Experiments with overlaps eliminated:  summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
So we will have to check whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimun as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

@@

In fact, Table \ref{tab:correlation}, which shows
the correlation between the number of IPs from one minute to the
next, presents a similar  behavior for all experiments but the last one: a
very high correlation indicating that the number of volunteers changes
slowly. The {\tt cachereboot} experiment correlation indicates a bigger
dynamism, with individuals joining and leaving the experiment all the
time. This might be circumstancial, but it also informs the fact
that the solution is found in less time and that users need less
contributions to find it.

@@

This future work holds many challenges which once again will go in two
different directions: gathering more users and improving the
efficiency of the algorithm run in an asynchronous way by the
users. In the first case, it will be interesting to make it more {\em
  social} by comparing local performance to other's contributions, and
showing it in a graph or being able to tweet advances or the fact that
the solution has been found. Anything that keeps the users running the
system and attracts new users will contribute to the speed of finding
the solution. And the algorithm will have to be improved, mainly by
making the population more diverse. Right now there is a rigid policy
of contribution to the pool: send the best individual, but this could
be relaxed. Some other policies could also be tested, as well as
running heterogenous algorithms in the same way that was done in
\cite{DBLP:journals/grid/ValdezTGVO15}. Diversity, as in any other
evolutionary algorith, seems to be the key, but there are many ways of
approaching it.

All these avenues of experimentation will be done openly following the
Open Science policy of our group, which, in fact, contributes to
establish trust and security between us and volunteers and is an
essential feature of the system. That is why this paper, as well as
the data and processing scripts, are published with a free license in GitHub at
\url{https://github.com/JJ/modeling-volunteer-computing}.

@@

This work has been supported in part by TIN2014-56494-C4-3-P (Spanish
Ministry of Economy and Competitivity), PROY-PP2015-06 (Plan Propio
2015 UGR) . We would also like to thank the anonymous reviewers of
previous versions of this paper who have really helped us to improve
this paper (and our work) with their suggestions. We are also grateful
to Anna S\'aez de Tejada for her help with the data processing
scripts. We are also grateful to {\tt @otisdriftwood} for his help
gathering users for the new experiments.

>>>>

\begin{abstract}
In a volunteer distributed computing system, users run a program on
their own machine to contribute to a common effort. If the program is
embedded in a web page, collaboration is easy, but also ephemeral. In
this paper, we analyze a volunteer evolutionary computing system called
NodIO, trying to find out a few rules that encourage volunteer
participation and how to make the most of their contributions to the
evolutionary algorithm. We will show different measures of
participation and contribution to the algorithm and how different
volunteer usage patterns and tweaks in the algorithm, such as restarting clients when solution has been found, contribute to
improvements and leveraging of these contributions. 
\end{abstract}

@@

The world wide web  provides not only a platform for content
distribution, but also an increasingly reliable and high-performance
operating system for running distributed applications. This is based in two
key things: the Javascript virtual machine every browser runs and a
simplified standard interface for interacting with servers.
% These are "things", not "facts". As facts would be... "every browser runs a Javascript virtual machine and that there is a simplified standard interface for interacting with servers (?)
% changed to things - JJ
Thus
creating a distributed computing experiment is just a matter of making
a JavaScript application interchange information with a server, called
REST. From the point of view of the programmer, this involves
relatively common skills and no special libraries, since that
interface is built in the browser, and a simple application that
responds to those requests on the server side; both involve just a few
dozens lines of code additionally to whatever business logic your
application has. But, more importantly and from the point of view of
the user, that application can be run by simply visiting a web page.

@@

Volunteer computing involves a user running a program voluntarily
and, as such, has been deployed in many different ways from the
beginning of the Internet, starting with the SETI@home framework for
processing extraterrestrial signals \cite{david-seti:home}. However
the dual facts of the introduction of JavaScript as a universal language for the
browser and the browser itself as an ubiquitous web and Internet client has
made this combination the most popular for volunteer computing
frameworks such as the one we are using here, and whose first version
was described in \cite{DBLP:conf/gecco/GuervosG15}.

@@

In general, those papers go no further than a proof of concept,
attempting to gauge how many users join the effort and also how long
would it take for a concrete amount of users to reach the solution. In
general, they do not try to assess the performance of the algorithm
itself in the precise conditions they find in a volunteer computing
system, although papers such as the one by Laredo et
al. \cite{churn08:ijhpsa}, using models, try to find out the resulting
performance when the users are changing continuously according to a
Weibull distribution. In the case of Klein and Spector \cite{unwitting-ec}, the algorithm
is actually run on the server, using the browser mainly for fitness
evaluation so no actual contribution to the evaluation of the
evolutionary algorithm itself is made.

@@

All experiments are announced in the same way: we use social networks,
mainly Twitter, to announce the start of a new experiment. LinkedIn,
Facebook, and also private groups in Telegram and WhatsApp have been
used for announcements too. We also encourage friends and followers to
put an original post for announcing it, instead of just sharing, in
the way allowed by the platform, the original post. During the running
of the experiments, users are engaged, questions answered, and all
kind of explanations given if requested. In general, announcements
have been made in Spanish, but in some cases English was used
too. This makes for a certainly unreliable experimentation framework,
but it is a realistic one. The steps in the design of the evolutionary
algorithm will be explained next.

@@

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=3>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Ranked and normalized number of contributions (PUTs) per IP,
  plotted in a log $y$ scale.\label{fig:zipf}}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP request) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that are slightly superior to the rest, specially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

%
\begin{table}
  \caption{Experiments with overlaps eliminated:  summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
So we will have to check whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

@@

In fact, Table \ref{tab:correlation}, which shows
the correlation between the number of IPs from one minute to the
next, presents a similar  behavior for all experiments but the last one: a
very high correlation indicating that the number of volunteers changes
slowly. The {\tt cachereboot} experiment correlation indicates a bigger
dynamism, with individuals joining and leaving the experiment all the
time. This might be circumstantial, but it also informs the fact
that the solution is found in less time and that users need less
contributions to find it.

@@

This future work holds many challenges which once again will go in two
different directions: gathering more users and improving the
efficiency of the algorithm run in an asynchronous way by the
users. In the first case, it will be interesting to make it more {\em
  social} by comparing local performance to other's contributions, and
showing it in a graph or being able to tweet advances or the fact that
the solution has been found. Anything that keeps the users running the
system and attracts new users will contribute to the speed of finding
the solution. And the algorithm will have to be improved, mainly by
making the population more diverse. Right now there is a rigid policy
of contribution to the pool: send the best individual, but this could
be relaxed. Some other policies could also be tested, as well as
running heterogeneous algorithms in the same way that was done in
\cite{DBLP:journals/grid/ValdezTGVO15}. Diversity, as in any other
evolutionary algorithm, seems to be the key, but there are many ways of
approaching it.

All these avenues of experimentation will be done openly following the
Open Science policy of our group, which, in fact, contributes to
establish trust and security between us and volunteers and is an
essential feature of the system. That is why this paper, as well as
the data and processing scripts, are published with a free license in GitHub at
\url{https://git.io/gecco-es-15}.

@@

This work has been supported in part by TIN2014-56494-C4-3-P (Spanish
Ministry of Economy and Competitivity), PROY-PP2015-06 (Plan Propio
2015 UGR). We are also grateful to {\tt @otisdriftwood} for his help
gathering users for the new experiments.
